#####################################################################################################
# File: Anstephensi_pop_PCA.rmd
#
# Created by Jeanne N. Samake in June 2022  
# For use in R
#
# Description: Principal component analysis of Anopheles stephensi populations based on independent 
# bi-allelic SNPs from reduced representation genetic data.
#
#NOTE: This script requires PLINK v1.9
#####################################################################################################


#(1) Generate PCA output using PLINK 
```{r }
system("./plink --bfile autosomal.plink.INDEP --maf 0.004 --geno 0.20 --pca --out autosomal.plink.INDEP.pca --noweb")
```

#Analyze and visualize PCA results 

#(2) Load tidyverse
```{r}
remove.packages("cli")
install.packages("cli")

install.packages("tidyverse", dependencies = TRUE)
library(tidyverse)
```

#(3) Upload PCA and Pop/Family ID files
```{r }
eigenValues <- read_delim("autosomal.plink.INDEP.pca.eigenval", delim = " ", col_names = F)
eigenVectors <- read_delim("autosomal.plink.INDEP.pca.eigenvec", delim = " ", col_names = F)
famDetails <- read.csv("samplePopId.csv", header = TRUE)

#Create sampleId column in eigenVector dataframe
eigenVectors$extractionID <- substr(eigenVectors$X1,1,5)

#Change column names of family details
colnames(famDetails) <- c("extractionID", "Pop.ID", "Sample.ID")

#Join eigenVector with sample data
newEigenVectors <- left_join(eigenVectors, famDetails, by = "extractionID")

#Remove unwanted columns in newEigenVectors dataframe
pcaEigenVectors <- newEigenVectors[c(24,2:22)]

```

#(4) Proportion of variation captured by each vector
```{r }
eigen_percent <- round((eigenValues / (sum(eigenValues))*100), 2)
```

#(5) PCA plot
```{r }
pcaEigenVectors$Pop.ID = factor(pcaEigenVectors$Pop.ID)
ggplot(data = pcaEigenVectors) +
  geom_point(mapping = aes(x = X3, y = X4, color = Pop.ID, shape = Pop.ID), size = 3, show.legend = TRUE ) +
  geom_hline(yintercept = 0, linetype="dotted") +
  geom_vline(xintercept = 0, linetype="dotted") +
  scale_shape_manual(values = 0:6) +
  labs(x = paste0("Principal component 1 (",eigen_percent[1,1]," %)"),
      y = paste0("Principal component 2 (",eigen_percent[2,1]," %)"),
      colour = "An. stephensi Population", shape = "An. stephensi Population") +
  theme_classic(base_size = 15)

 ggsave("pca_AnsteInd.png", dpi=600)
```

#(6)Redo PCA analysis with subregion groups
```{r }
eigenValues <- read_delim("autosomal.plink.INDEP.pca.eigenval", delim = " ", col_names = F)
eigenVectors <- read_delim("autosomal.plink.INDEP.pca.eigenvec", delim = " ", col_names = F)
famDetails2 <- read.csv("samplePopId2.csv", header = TRUE)

#Change column names of subgroup/family details
colnames(famDetails2) <- c("extractionID", "Pop.ID", "Sample.ID", "Group")

#Join eigenVector with sample data
newEigenVectors2 <- left_join(eigenVectors, famDetails2, by = "extractionID")

#Remove unwanted columns in newEigenVectors dataframe
pcaEigenVectors2 <- newEigenVectors2[c(24,2:22,26)]

```

#(7) PCA with subgroups plot
```{r}
pcaEigenVectors2$Pop.ID = factor(pcaEigenVectors2$Pop.ID)
Plot2 <- ggplot(data = pcaEigenVectors2) +
  geom_point(mapping = aes(x = X3, y = X4, color = Group, shape = Group), size = 3, show.legend = TRUE ) +
  geom_hline(yintercept = 0, linetype="dotted") +
  geom_vline(xintercept = 0, linetype="dotted") + #
  stat_ellipse(aes(x = X3, y = X4, color = Group), type = "norm") +
  labs(x = paste0("Principal component 1 (",eigen_percent[1,1]," %)"),
      y = paste0("Principal component 2 (",eigen_percent[2,1]," %)"),
      colour = "An. stephensi Pop", shape = "An. stephensi Pop") +
  theme_classic(base_size = 15)

Plot2 + scale_colour_discrete(name  ="An. stephensi Population",
                            breaks=c("Subgroup 1", "Subgroup 2"),
                            labels=c("Subgroup 1", "Subgroup 2")) +
      scale_shape_discrete(name  ="An. stephensi Population",
                           breaks=c("Subgroup 1", "Subgroup 2"),
                           labels=c("Subgroup 1", "Subgroup 2"))

ggsave("pca_Anste_subgroups.png", dpi=600)
```


